Realizing temporal graphs from fastest travel times

### 2023-09-22 STACS2023/nina_shortversion.tex

\subsubsection{First part of the FPT algorithm}
We now present the first part of our FPT algorithm.
Let $D$ be the input matrix of \deltaExact\, and let $G$ be its underlying graph, on $n$ vertices and $m$ edges.
%\subsubsection*{Preprocessing of the input}
From the underlying graph $G$ of $D$ we first create a graph $G'$ by
iteratively removing vertices of degree one from $G$,
and denote with 
$
Z = V(G) \setminus V(G')
$, the set of removed vertices.
%Then we determine a minimum feedback edge set $F$ of $G'$.
%by finding a spanning tree $T$ of $G'$ and set $F = E(G) \setminus E(T)$. 
%Note that $F$ is also a minimum feedback edge set of $G$.
Then we determine sets $U$, of \emph{vertices of interest}, and $U^*$ of the neighbors of vertices of interest, in the following way.
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subset V(G')$ be the set of leaves in the spanning tree $T$, $V_2 \subset V(G')$ be the set of vertices of degree two in $T$, that are incident to at least one edge in $F$, 
and let $V_3 \subset V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ is incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with 
\[U = V_1 \cup V_2 \cup V_3\]
the set of vertices of interest. It follows that $|U| \leq 4k$.
We set $U^*$ to be the set of vertices in $V(G') \setminus U$ that are neighbors of vertices in $U$, \ie 
\[U^* = \{v \in V(G') \setminus U  :  u \in U, v \in N(u)\}.\]
Again, using the tree structure, we get that for any $u \in U$ its neighborhood is of size $|N(u)| \in O(k)$, since every neighbor of $u$ is the first vertex of a (unique) path to another vertex in $U$.
It follows that $|U^*| \in O(k^2)$.
From the construction of $Z$ (by iteratively removing vertices of degree one from $G$)
it follows that $Z$ consists of disjoint trees $T_1, T_2, \dots$.
For a tree $T_i$ we denote with $u_i$ the vertex in $G'$ that is a neighbor of a vertex in $T_i$, 
and call it a \emph{clip vertex of the tree $T_i$}.
It follows that there can be many different trees $T_i$ that are incident to the same clip vertex $u_i \in V(G')$,
but each tree $T_i$ is incident to exactly one clip vertex $u_i \in V(G')$.
Since $u_i$ is the only vertex connecting all of the trees $T_i$ incident to it,
from now on we assume that a tree $T_{u_i}$ in $Z$ is a union of trees on vertices from $V(G) \setminus V(G')$,
that are clipped at the same vertex $u_i \in V(G')$.
For each of the trees $T_{u_i}$ in $Z$, we select one vertex $r_i$, that is a neighbor of the clip vertex $u_i$,
and call it \emph{a representative vertex of the tree $T_{u_i}$}.
We now define as $Z^*$ the set of representatives $r_i$ of trees $T_i \in Z$,
where the clip vertex $v_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in T_i, \text{ where $T_i \in Z$, the clip vertex $u_i$ of $T_i$ is in $U$, and $r_i u_i \in E(G)$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
Note that determining sets $U$, $U^*$, and $Z^*$ takes linear time.

%Let us remember the following basic property of the problem \deltaExact.
Recall that a labeling $\lambda$ of $G$ satisfies $D$ if the duration of a fastest temporal path from vertex $v_i$ to $v_j$ equals $D_{v_i, v_j}$.
%, and no other temporal path from $v_i$ to $v_j$ is faster.
In order to find a labeling that satisfies this property we iterate through all pairs of vertices $v_i, v_j \in V(G)$ and 
first guess the underlying path $P$ of a fastest temporal path from $v_i$ to $v_j$, which results in one equality constraint for the labels on the path $P$. 
For all other temporal paths from $v_i$ to $v_j$ we know that they cannot be faster, so we introduce inequality constraints for them.
This results in producing $poly(n,m)$ constraints. 
It is important to point out that the number of constraint and number of variables are not related, while we introduced all of the necessary constraints, we manage to create only a small number of variables (upper-bounded by some function in $k$).
%Note that we have to do this while keeping the total number of variables upper-bounded by some function in $k$, therefore we do not 
%
For an easier understanding and the analysis of the algorithm we give the following definition.
\begin{definition}\label{def:segments}
	Let $U \subseteq V(G')$ be a set of vertices of interest and let $u,v \in U$.
	A path $P = (u=v_1,v_2, \dots, v_p = v)$  in graph $G'$, where all inner vertices are not in $U$, \ie $v_i \notin U$ for all $i \in \{ 2, 3, \dots, p-1\}$,
	is called a \emph{segment} from $u$ to $v$. We denote it as $S_{u,v}$.
\end{definition}
%From the definition it follows, that there are $O(k^2)$ segments in $G'$. \todo{True but does not follow directly just from the definition}
Note from \cref{def:segments} we get that $S_{u,v} \neq S_{v,u}$. 
Observe that a temporal path in $G'$ between two vertices of interest is either a segment, or consists of a sequence of some segments. Furthermore, since we have at most $4k$ interesting vertices in $G'$, we can deduce the following important result.
\begin{corollary}\label{obs:FPT-k2segments}
	There are at most $O(k^2)$ segments in $G'$.
\end{corollary}
%In the analysis of the algorithm we rely heavily on the fact that there are $O(k^2)$ segments in $G'$.
%, which we prove later on (see~\cref{obs:FPT-k2segments}).
%For a pair $u,v$ of vertices in $U$, there are $O(k!)=O(k^k)$ possible paths in $G'$ between them.

%\subsubsection{Guessing necessary structures \label{sec:FPT-guessing}}
Once the sets $U, U^*$ and $Z^*$ are determined, we are ready to start guessing  the necessary structures.
Note that whenever we say that we guess the fastest temporal path between two vertices, we mean that we guess the underlying path of a representative fastest temporal path between those two vertices.

@@

\noindent

@@

\todo{Argue why each (some) of the guesses are important}

>>>>

\subsubsection{First part of the FPT algorithm}
We now present the first part of our FPT algorithm.
Let $D$ be the input matrix of \deltaExact\, and let $G$ be its underlying graph, on $n$ vertices and $m$ edges.
%\subsubsection*{Preprocessing of the input}
From the underlying graph $G$ of $D$ we first create a graph $G'$ by
iteratively removing vertices of degree one from $G$,
and denote with 
$
Z = V(G) \setminus V(G')
$, the set of removed vertices.
%Then we determine a minimum feedback edge set $F$ of $G'$.
%by finding a spanning tree $T$ of $G'$ and set $F = E(G) \setminus E(T)$. 
%Note that $F$ is also a minimum feedback edge set of $G$.
Then we determine sets $U$, of \emph{vertices of interest}, and $U^*$ of the neighbors of vertices of interest, in the following way.
Let $T$ be a spanning tree of $G'$, with $F$ being the corresponding feedback edge set of $G'$.
Let $V_1 \subset V(G')$ be the set of leaves in the spanning tree $T$, $V_2 \subset V(G')$ be the set of vertices of degree two in $T$, that are incident to at least one edge in $F$, 
and let $V_3 \subset V(G')$ be the set of vertices of degree at least $3$ in $T$. 
Then $|V_1| + |V_2| \leq 2k$, since every leaf in $T$ and every vertex in $V_2$ is incident to at least one edge in $F$,
and $|V_3| \leq |V_1|$ by the properties of trees.
We denote with 
\[U = V_1 \cup V_2 \cup V_3\]
the set of vertices of interest. It follows that $|U| \leq 4k$.
We set $U^*$ to be the set of vertices in $V(G') \setminus U$ that are neighbors of vertices in $U$, \ie 
\[U^* = \{v \in V(G') \setminus U  :  u \in U, v \in N(u)\}.\]
Again, using the tree structure, we get that for any $u \in U$ its neighborhood is of size $|N(u)| \in O(k)$, since every neighbor of $u$ is the first vertex of a (unique) path to another vertex in $U$.
It follows that $|U^*| \in O(k^2)$.
From the construction of $Z$ (by iteratively removing vertices of degree one from $G$)
it follows that $Z$ consists of disjoint trees $T_1, T_2, \dots$.
For a tree $T_i$ we denote with $u_i$ the vertex in $G'$ that is a neighbor of a vertex in $T_i$, 
and call it a \emph{clip vertex of the tree $T_i$}.
It follows that there can be many different trees $T_i$ that are incident to the same clip vertex $u_i \in V(G')$,
but each tree $T_i$ is incident to exactly one clip vertex $u_i \in V(G')$.
Since $u_i$ is the only vertex connecting all of the trees $T_i$ incident to it,
from now on we assume that a tree $T_{u_i}$ in $Z$ is a union of trees on vertices from $V(G) \setminus V(G')$,
that are clipped at the same vertex $u_i \in V(G')$.
For each of the trees $T_{u_i}$ in $Z$, we select one vertex $r_i$, that is a neighbor of the clip vertex $u_i$,
and call it \emph{a representative vertex of the tree $T_{u_i}$}.
We now define as $Z^*$ the set of representatives $r_i$ of trees $T_i \in Z$,
where the clip vertex $v_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in T_i, \text{ where $T_i \in Z$, the clip vertex $u_i$ of $T_i$ is in $U$, and $r_i u_i \in E(G)$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
To solve \deltaExact\ correctly it is necessary that for every pair of vertices $v_i, v_j \in V(G)$ we
\begin{itemize}
    \item determine the underlying path $P$ of the fastest temporal path from $v_i$ to $v_j$,
for which we introduce the equality constraint (\ie $d(P) = D_{v_i, v_j}$), and %in the ILP
\item make sure that every other path $P'$ from $v_i$ to $v_j$ is not faster than $P$ (\ie $d(P') \geq D_{v_i, v_j}$). %in the ILP
\end{itemize}
One could easily achieve this by ``guessing'' the fastest paths among each pair of vertices, but this would result in too many guesses. 
For problem to remain feasible, one should create only a small number of guesses.
With this in mind we create the following guesses.
%It's important to mention that whenever we say that we guess the fastest temporal path between two vertices, we mean that we guess the underlying path of a representative fastest temporal path between those two vertices.

@@

Unfortunately the above guesses do not provide us with enough information to be able to determine all fastest paths. For example consider the case depicted in the FIGURE TODO. 
Therefore we introduce some more guesses. To do this we have to first define the following.
\begin{definition}\label{def:segments}
	Let $U \subseteq V(G')$ be a set of vertices of interest and let $u,v \in U$.
	A path $P = (u=v_1,v_2, \dots, v_p = v)$  in graph $G'$, where all inner vertices are not in $U$, \ie $v_i \notin U$ for all $i \in \{ 2, 3, \dots, p-1\}$,
	is called a \emph{segment} from $u$ to $v$. We denote it as $S_{u,v}$.
\end{definition}
Note, from \cref{def:segments} we get that $S_{u,v} \neq S_{v,u}$. 
Observe that a temporal path in $G'$ between two vertices of interest is either a segment, or consists of a sequence of some segments. Furthermore, since we have at most $4k$ interesting vertices in $G'$, we can deduce the following important result.
\begin{corollary}\label{obs:FPT-k2segments}
	There are at most $O(k^2)$ segments in $G'$.
\end{corollary}
We now continue with our guessing.

@@
